---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# Household Latrine User Survey – Dowa & Thyolo Malawi (2020)

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

This dataset was collected in October 2020 from Dowa and Thyolo Districts, Malawi,
using the mWater survey platform. It captures detailed information on
household sanitation practices and latrine usage across both rural and
peri-urban communities. The survey was conducted by trained enumerators
working with BASEflow Malawi as part of efforts to document and improve
Water, Sanitation, and Hygiene (WASH) infrastructure and behavior in the
region.

**Potential Use Cases**

This data is a valuable resource for multiple stakeholders, including:

1.  Local Government & Planners: To inform sanitation infrastructure
    planning and budgeting in rural and peri-urban Malawi.

2.  Public Health Practitioners: To study the relationship between
    latrine use and hygiene-related health outcomes.

3.  WASH Sector NGOs: To identify service delivery gaps, monitor impact,
    or design targeted behavior change programs.

4.  Researchers & Academics: For sanitation behavior studies,
    infrastructure sustainability, and social acceptability of different
    latrine models.

5.  Donors & Development Partners: To assess community sanitation needs,
    target investments, and support SDG 6 (Clean Water and Sanitation).

## Installation

You can install the development version of dowalatrinesurvey from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/dowalatrinesurvey")
```

```{r}
## Run the following code in console if you dont have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/dowalatrinesurvey/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to Latrine Household Data for Dowa and Thyolo Districts
in Malawi for the year 2020.

```{r}
library(dowalatrinesurvey)
```

### dowalatrinesurvey

The dataset `dowalatrinesurvey` contains `r nrow(dowalatrinesurvey)`
observations and `r ncol(dowalatrinesurvey)` variables

```{r}
dowalatrinesurvey |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "dowalatrinesurvey.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Visualizations

```{r}
library(dowalatrinesurvey)

# Visualization 1 Spatial Distribution Map by Community Type
# Purpose is to show how households with different community_type (Rural vs Peri-Urban) are spatially distributed across your study area.

library(dplyr)
library(ggplot2)
library(maps)
library(stringr)
# Clean community_type in dowalatrinesurvey
dowalatrinesurvey <- dowalatrinesurvey %>% mutate(community_type = trimws(community_type))

# Get Malawi map data from maps package
malawi_map <- map_data("world", region = "Malawi")

# Plot using dowalatrinesurvey
ggplot() +
  geom_polygon(
    data = malawi_map,
    aes(x = long, y = lat, group = group),
    fill = "gray90",
    color = "black"
  ) +
  geom_point(
    data = dowalatrinesurvey,
    aes(x = longitude, y = latitude, color = community_type),
    size = 3, alpha = 0.7
  ) +
  coord_fixed(1.3) +  # fix aspect ratio for lat/lon
  labs(
    title = "Spatial Distribution of Surveyed Households by Community Type",
    x = "Longitude",
    y = "Latitude",
    color = "Community Type"
  ) +
  theme_minimal()

# Visualization 2: Grouped bar chart showing count of latrine satisfaction by community type
# visually compare the levels of satisfaction with latrines between different types of communities
# (e.g., Rural vs Peri-Urban)

# Clean and factor satisfaction variable
data <- dowalatrinesurvey %>%
  mutate(
    satisfaction_clean = str_extract(satisfaction_with_latrine, "^[0-9]"),
    satisfaction_clean = factor(satisfaction_clean, levels = c("1", "2", "3"),
                                labels = c("Not Satisfied", "Neutral", "Satisfied"))
  ) %>%
  filter(!is.na(satisfaction_clean))  # Remove NA

# Summarize counts
satisfaction_summary <- data %>%
  group_by(community_type, satisfaction_clean) %>%
  summarise(count = n(), .groups = "drop")

# Plot grouped bar chart (counts)
ggplot(satisfaction_summary, aes(x = community_type, y = count, fill = satisfaction_clean)) +
  geom_bar(stat = "identity", position = "dodge") +  # side-by-side bars
  labs(
    title = "Latrine Satisfaction by Community Type",
    x = "Community Type",
    y = "Count",
    fill = "Satisfaction Level"
  ) +
  theme_minimal()

```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("dowalatrinesurvey")
```
